Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-03-12 manuscript/manuscript.tex
Updated keywords
manuscript_0311

\abstract{Good advice improves decision quality but often requires
delivering unpleasant truths that may disappoint advisees. Across three
pre-registered and incentivized experiments involving real
adviser-advisee interactions (N = 3,900), we show that advisers
prioritize avoiding disappointment at the expense of accuracy and their
own earnings. In Study 1, advisers financially rewarded for accuracy
still tailor recommendations to aspirational goals expressed by
advisees, resulting in worse advice. When incentivized to be liked,
advisers provide even more flattering advice, and advisees reward this
by rating these advisers as more likable, despite the advice being less
honest and less accurate (Study 2). The desire to avoid disappointment
may lead to inequities if there are differences in expectations across
social groups. In Study 3, we examine a setting in which men expect to
perform better than women. We show that advisers take into account these
expectations, leading to systematically different advice for men and
women even when their gender is unknown to advisers. Advisers' efforts
to avoid disappointment may thus contribute to systematic gender
disparities in advice, with implications for downstream
decision-making.}

\keywords{Gender difference, Advice, Interpresonal relationship}

@@

Career trajectories for men and women often differ substantially, even
among those with similar qualifications. One source of such differences
may be the advice that people receive. Prior work has documented that
men receive more aspirational advice and women more risk-averse advice,
which may explain why women are less likely to apply for and ultimately
obtain more rewarding positions (Kanze et al., 2018). Research on gender
discrimination has focused on the role of unconscious bias (Banaji \&
Greenwald, 1995; Nosek et al., 2009), but interventions that try to
reduce this bias have been largely unsuccessful (Chang et al., 2019;
Paluck \& Green, 2009). Recent work has proposed cognitive, rather than
unconscious, biases that can lead to gender discrimination through the
formation of false beliefs (Hagmann et al., 2024). Here, we propose a
novel source of differences in advice by drawing on recent work on
belief-based utility, expectation-based reference points (Kőszegi \&
Rabin, 2006, 2009), and the hedonic consequences of information (Golman
et al., 2017).

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
(Gino \& Moore, 2007; Harvey \& Fischer, 1997; Soll \& Larrick, 2009).
Its fundamental premise is to improve outcomes by enabling
decision-makers to learn from others' expertise. However, while honest
advice can lead to better decisions, it can also lead to disappointment
when it suggests a desired outcome is unlikely to materialize. Advisers
may be motivated to offer flattering recommendations that align with
advisees' hopes, sparing them the disappointment of a more sobering
forecast.

Advice with the potential for disappointment appears in diverse
organizational settings: supervisors deliver feedback on employee
performance, lawyers convey likely outcomes of litigation to their
clients, and academics offer thoughts on colleagues' manuscripts.
Ideally, honest information would help people identify their strengths
and weaknesses, make informed decisions about whether to pursue legal
cases, and strengthen papers prior to submission to academic journals,
thus improving long-term outcomes. In practice, however, interpersonal
considerations may prevent candor. When someone's expectations are (too)
high, an adviser might hesitate to deliver information that undermines
it---particularly if doing so risks negative interpersonal
repercussions.

While much of the literature on advice assumes that advisers aim to
provide the most helpful advice, thier motivations are often more
complex (Dalal \& Bonaccio, 2010; Rader et al., 2017; Yaniv \&
Choshen-Hillel, 2012). Advisers are not only concerned with the accuracy
and usefulness of their advice but also with their relationships with
advisees. Interpersonal concerns---such as maintaining a positive image,
avoiding conflict, or preserving social harmony---can shape the way
advice is delivered (Chentsova-Dutton \& Vaughn, 2012; Schaerer et al.,
2018; Van Swol, 2009). In particular, advisers may strategically soften
their messages or withhold unfavorable information to manage how they
are perceived or to prevent discomfort for both themselves and the
advisee (Levine et al., 2020; Sniezek \& Van Swol, 2001). These
interpersonal dynamics suggest that advice is not always optimized for
the recipient's best interest but is instead filtered through social and
psychological considerations.

Just as advisers may hesitate to provide unwelcome truths, advisees
themselves are often resistant to hearing them. People often go to great
lengths to avoid unpleasant information, deriving belief-based utility
from maintaining favorable views of themselves (Golman et al., 2017; Ho
et al., 2021; Loewenstein \& Molnar, 2018). Moreover, they often prefer
advisers who withhold bad news from them (Shalvi et al., 2019) and
punish those who communicate bad news (John et al., 2019). Advisers,
anticipating these preferences and consequences face a dilemma: Should
they provide the accurate but potentially painful truth, or shade their
recommendations to preserve the advisee's mood and the adviser's own
standing? Advisers who fear they will be blamed or disliked for
undermining someone's confidence may reasonably avoid candid feedback,
particularly if they don't incur a cost when the advisee makes a
mistake.

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their feedback to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.\footnote{In some cases, however,
  repeated interaction might also increase incentives for honest
  feedback. Someone who is known to persistently give overoptimistic
  advice may be viewed as less trustworthy in the long run.}

In Study 1, we nudge advisees into a preference for competing against
either a group of top-performers or a group of low-performers, through
the use of a default option. Advisers who observe the actual performance
of advisees and therefore have the relevant information needed to make a
recommendation nonetheless take into account the initial preference when
recommending which group they should compete against. In Study 2, we ask
advisees to upload photos of themselves and advisers rank them on their
attractiveness. Advisers then recommend which rank the advisee should
bet on, receiving an incentive either for the advisee's accuracy or for
being evaluated favorably. Both groups of advisers recommend a rank that
is more attractive than what they themselves have evaluated the advisee,
and those incentivized for likability further inflate the ranking.
Advisees evaluate advisers more favorably when they recommend betting on
a more attractive rank, including viewing them as more trustworthy.

@@

We first examine whether advisers take into account the expectations of
advisees when providing a recommendation. A group of participants
(``Advisees'') complete a quiz consisting of ten questions that draw on
ego-relevant domains. They then express a non-binding preference to
compete against a group of high performers or low performers on the same
task, based on their past performance. We nudge participants towards
picking either the low or the high performer group by selecting one of
the two by default. A second group of participants (``Advisers'')
observe Advisees' score on the quiz and, in one treatment, also which
group they had selected. We predicted that Advisees who were nudged
toward the high performer group would be more likely to receive advice
to compete against this high group when Advisers observed their
non-binding choice.

@@

@@

{\caption{{Advice to compete against the High Performer Group in Study
1. Displaying the non-binding choice of those for whom the Low Performer
Group was selected by default makes it less likely that they are advised
to compete against the High Performer Group (Column 1). Column 2
restricts the analysis to advisers who observe advisees' initial
non-binding choice and controls for advisees' score on the quiz.
Standard errors are clustered at the adviser
level.}{\label{tbl-study1regs}}}\vspace{0pt}
}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]},
column{3}={}{halign=c,},
column{1}={}{halign=l,},
cell{2}{2}={}{halign=c,},
cell{3}{2}={}{halign=c,},
cell{4}{2}={}{halign=c,},
cell{5}{2}={}{halign=c,},
cell{6}{2}={}{halign=c,},
cell{7}{2}={}{halign=c,},
cell{8}{2}={}{halign=c,},
cell{9}{2}={}{halign=c,},
cell{10}{2}={}{halign=c,},
cell{11}{2}={}{halign=c,},
cell{12}{2}={}{halign=c,},
cell{13}{2}={}{halign=c,},
cell{1}{2}={c=2,}{halign=c, halign=c,},
hline{13}={1,2,3}{solid, black, 0.05em},
}                     %% tabularray inner close
\toprule
& Advice to Compete Against High Performers &  \\ \cmidrule[lr]{2-3}
& All Advice & Expectations Shown \\ \midrule %% TinyTableHeader
High Default               & \num{0.011}    & \num{0.025}*    \\
& (\num{0.013})  & (\num{0.011})   \\
Expectation Shown          & \num{-0.041}*  &                  \\
& (\num{0.016})  &                  \\
Expectation x High Default & \num{0.037}+   &                  \\
& (\num{0.019})  &                  \\
Score                      &                 & \num{0.168}***  \\
&                 & (\num{0.004})   \\
Constant                   & \num{0.389}*** & \num{-0.503}*** \\
& (\num{0.012})  & (\num{0.020})   \\
N                          & \num{10000}    & \num{5000}      \\
\bottomrule
\end{talltblr}

@@

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument that
reflect higher expectations. We show linear probability models for the
decision of the Advisers, which is equal to 1 if they recommend
competing against the High Performer Group and 0 if they recommend the
Low Performer Group. In Column 1 of Table~\ref{tbl-study1regs}, we show
our main specificiation with the experimental assignment for Advisees
(High Performer Default vs.~Low Performer Default) and Advisers
(Expectation Shown vs.~Expectation Hidden), as well as their
interaction. Because each adviser made ten recommendations, we cluster
standard errors at the adviser level. We find that displaying
expectations of those in the Low Performer Default reduces
recommendations to compete against the Top group by 4 percentage points,
or about 10\% relative to when expectations are hidden. This decrease is
marginally smaller for advisees with a default choice to compete against
the High Performer Group (nudged to have high expectations, though not
necessarily holding them), and whose advice did not differ based on
whether their expectations were shown or hidden. We reported a linear
probability model on advising to compete against the High Performer
group for advisers in the ``Expectations'' treatment, controlling for
the true performance of the advisee in Column 2 of
Table~\ref{tbl-study1regs}. We observe a significant main effect of the
default treatment, where a high default increased the likelihood of
receiving flattering advice. It is worth noting that, since only 65 \%
of advisees in the high default group actually stuck with their default
choice, this represents a conservative test of the effect of high
expectations on the likelihood of receiving flattering advice.

@@

\caption{\label{fig-study1Advice}Advice given in Study 1. Advisees who
were nudged to express high expectations were more likely to receive
advice to compete against the High Performer Group.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study1Advice-1.pdf}}

@@

When an adviser knows an outcome that an advisee can only estimate,
learning this estimate should not change the advice that is given.
However, in line with our argument that advisers take into account the
belief utility of the advisee and prefer to avoid disappointing them, we
find that providing information about expectations makes a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive flattering advice, encouraging
them to pursue that option. This occurs even though advisers are
incentivized to give good advice.

We next examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We
do this by manipulating the incentives for advisers, who receive a bonus
either based on how they are evaluated by an advisee, or a bonus for the
advisee's accuracy.

\caption{\label{fig-study2advice}Participants in Study 2 give advice to
the person they rank 7th most attractive out of 10. When incentivized
for accuracy, the majority recommends betting on that rank. However,
when incentivized for likeability, participants inflate their advice to
flatter the recipient as more attractive.}

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study2advice-1.pdf}}

Next, we turn our attention to the advisers (see
Figure~\ref{fig-study2advice}). In the Accuracy condition, those
uploading selfies are on average advised to bet on rank 6.19. Notably,
this is significantly more attractive than the 7th rank those advisers
had themselves guessed just on the prior screen (\(t(234) = -8.82\),
\(p < .001\)). This suggests that even when incentivized for accuracy,
participants offer flattering advice.\footnote{This shading could be due
  to concerns of avoiding disappointment. However, it could also be that
  advisers are uncertain about the rankings they have given and make a
  recommendation that combines their own belief with a uniform prior.
  Therefore, our analyses focus on the difference between the two
  conditions.} Importantly, and as predicted, we find that advisers in
the Likeable treatment recommend betting on a lower rank, communicating
that they think the participant in the selfie is more attractive (5.38,
\(t(470) = 5.44\), \(p < .001\)). This suggests that advisers infer
flattering someone with pleasant advice would make the adviser appear
more likeable. The distribution shown in Figure~\ref{fig-study2advice}
shows that participants do not simply tell participants that they are
the most attractive person in the group. They may infer that flattering
advise needs to be somewhat realistic to be believable. We return to
this in the general discussion.

>>>>

\abstract{Advice that improves decision-making does not always feel
good. While advice has the potential to improve decision-making by
helping advisees more accurately assess feasibility and predict
potential outcomes, it can also lead to disappointment when it
challenges the advisee's overly optimistic expectations. Across three
pre-registered and incentivized experiments involving real
adviser-advisee interactions (N = 3,900), we show that advisers
prioritize avoiding disappointment at the expense of accuracy and their
own earnings. In study 1, advisers use flattering advice to create more
positive impressions, and advisees reward this by rating these advisers
as more likable, despite the advice being less honest and less accurate.
Even advisers incentivised for accuracy still tailor recommendations to
aspirational goals expressed by advisees, resulting in financial loss
for them and the advisees (Study 2). The concern for interpersonal
consequences may lead to inequities if there are differences in
expectations across social groups. In Study 3, we examine a setting in
which men expect to perform better than women. We show that advisers
take into account these expectations, leading to systematically
different advice for men and women even when their gender is unknown to
advisers. Advisers' efforts to avoid disappointment may thus contribute
to systematic gender disparities in advice, with implications for
downstream decision-making.}

\keywords{Advice, Interpresonal relationship, Gender difference}

@@

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
(Gino \& Moore, 2007; Harvey \& Fischer, 1997; Soll \& Larrick, 2009).
While others can provide information in the form of opinions (Trouche et
al., 2018), feedback (Kluger \& DeNisi, 1996), or personal observations
(Yoon et al., 2021) to influence the decision-maker, this project
focuses specifically on advice---defined as ``information that is
perceived as intended to help an advisee deal with their issue of
deliberation'' (Milyavsky \& Gvili, 2024). Advice is distinct from other
forms of input because its primary purpose is to improve decision-making
by leveraging others' expertise.

Advice has the power to enhance decisions, yet it also carries the risk
of disappointment. For example, a career consultant might advise an
employee to enroll in a management program to strengthen their skill set
before competing for a leadership role. Ideally, such honest advice
helps the employee identify areas for personal growth. In practice,
however, the employee might feel overlooked and frustrated that their
ambitions are not immediately supported. As a result, advisers may be
inclined to offer flattering recommendations that align with---or even
exceed---advisees' hopes, such as encouraging the employee to pursue the
leadership role despite concerns about their readiness. In this way,
interpersonal dynamics may discourage full candor in giving advice.

The objective of this research is to examine the dilemma advisers face:
Should they provide accurate but potentially painful advice, or soften
their recommendations to preserve the advisee's mood and protect their
own standing? Given that advisees tend to prefer advisers who withhold
bad news (Shalvi et al., 2019) nd penalize those who deliver it (John et
al., 2019), advisers may reasonably avoid full candor---especially when
they bear no cost for the advisee's mistakes. Specifically, when
someone's expectations are (too) high, an adviser may hesitate to
provide advice that challenges them, particularly if doing so risks
straining the relationship or causing negative interpersonal
consequences.

Our research contributes to the growing literature on advisers' motives
and intentions in advice-giving. This line of work examines how
advisers' perceived dishonesty or helping intentions influence advice
utilization (Haran \& Shalvi, 2020; Mackinger et al., 2017; Milyavsky \&
Gvili, 2024; Van Swol, 2009). By examining advice within its
interactional and communicative context, we extend this research by
incorporating another dimension of advisers' motives---interpersonal
concerns. We demonstrate that advisers adjust their recommendations to
avoid hurting the advisee's feelings and self-image. Second, we
contribute to the belief-based utility literature, which suggests that
individuals derive satisfaction from their beliefs about the world and
themselves (Loewenstein \& Molnar, 2018). We show that advisers also
place value on preserving advisees' self-perceptions, often tailoring
their advice to align with the advisee's expectations. Furthermore, we
reveal that this advice-giving strategy has unintended consequences---it
transforms the confidence gap {[}where women tend to be less confident
than men; Lundeberg et al. (1994){]} into an advice gap, in which women
are less likely than men to receive advice to aim higher, thereby
reinforcing existing gender inequalities.

\subsection{The Motives Behind
Advice-Giving}\label{the-motives-behind-advice-giving}

While much of the literature on advice assumes that advisers aim to
provide the most helpful advice, their motivations are often more
complex (Dalal \& Bonaccio, 2010; Rader et al., 2017; Yaniv \&
Choshen-Hillel, 2012). Other than the accuracy and usefulness of the
advice, interpersonal concerns---such as maintaining a positive image,
avoiding conflict, or preserving social harmony---can shape what advice
is delivered and how (Levine et al., 2020; Sniezek \& Van Swol, 2001).
For example, research has shown how people often avoid providing
negative feedback directly, opting instead for ambiguity or indirect
phrasing to minimize discomfort (Levine et al., 2020). These
interpersonal dynamics suggest that advice is not always optimized for
the recipient's best interest but is instead filtered through social and
psychological considerations.

In particular, advisers may strategically soften their messages or
withhold unfavorable evaluations to manage how they are perceived or to
prevent discomfort for both themselves and the advisee (Chentsova-Dutton
\& Vaughn, 2012; Schaerer et al., 2018; Van Swol, 2009). When advisers
anticipate disagreement or resistance, they may soften their
recommendations or frame them more diplomatically, sometimes at the
expense of accuracy. This suggests that advice is not merely a means of
sharing information but also a tool for managing social interactions and
reducing potential friction (Van Swol, 2009). These interpersonal
considerations become even more pronounced when there is a power
imbalance between the adviser and advisee, as lower-power advisers may
be more hesitant to offer direct or critical guidance to avoid negative
repercussions (Schaerer et al., 2018).

Just as advisers may hesitate to provide unwelcome truths, advisees
themselves are often resistant to hearing them. People often go to great
lengths to avoid unpleasant information, deriving belief-based utility
from maintaining favorable views of themselves (Golman et al., 2017; Ho
et al., 2021; Loewenstein \& Molnar, 2018). Moreover, they often prefer
advisers who withhold bad news from them (Shalvi et al., 2019) and
punish those who communicate bad news (John et al., 2019). Advisers who
fear they will be blamed or disliked for undermining someone's
confidence may reasonably avoid candid advice, particularly if they
don't incur a cost when the advisee makes a mistake.

Providing flattering advice, however, should not be seen as maliciously
misleading the advisee. Rather than being irrational, it tends to be
more optimistic, and this optimism is not necessarily misplaced. Often,
the advisee possesses insights about themselves---such as their
motivation and past experiences---that the adviser may overlook or have
no access to. Take, for example, an employee deciding whether to compete
for a leadership role. While the adviser might focus on the employee's
lack of management knowledge, the employee may have strong leadership
characteristics, excellent communication skills, and persuasive
abilities. In this way, flattering advice can encourage the advisee to
set ambitious goals, fostering a self-fulfilling prophecy where
confidence drives success (Jussim, 1986; Locke \& Latham, 2006).

Of course, if the advice is overly inflated, it may become unhelpful.
However, this is unlikely to happen, as advisers typically aim to boost
the advisee's confidence while maintaining credibility. Suggesting an
unrealistic option---such as encouraging an entry-level employee to
compete for a CEO position---would come across as insincere or even
mocking rather than flattering.

Beyond protecting the advisee's beliefs and fostering their confidence
and motivation, advisers may also offer flattering advice to shield
themselves from blame if the ultimate decision leads to a poor outcome.
Although both the adviser and advisee recognize the inherent uncertainty
between the advice, the decision, and the final outcome, people tend to
attribute poor outcomes to the adviser (Palmeira et al., 2015; Weiner,
1985). However, research suggests that people differentiate between
errors made intentionally---such as those driven by insufficient
knowledge or deliberate bias---and those made unintentionally, with
harsher penalties imposed on advisers perceived to have deliberately
undermined the decision (Haran \& Shalvi, 2020). Flattering advice, even
when ultimately unhelpful, may reduce the likelihood of blame because
advisees are less likely to suspect ill intent from an adviser who
appears supportive. Advisees may instead interpret the poor outcome as
an unintended error rather than deliberate misguidance, particularly
because they view the adviser positively and believe the adviser
genuinely holds them in high regard (Haran \& Shalvi, 2020; Palmeira et
al., 2015). Even when advisees recognize that flattering advice was
intended to boost their confidence, research shows that people tend to
appreciate such misplaced good intentions, as they strengthen social
bonds by making the gift-giver feel socially closer (Zhang \& Epley,
2012).

Advisers may choose to inflate their advice for different reasons,
depending on their relationship with the advisee and the stakes of the
decision. For instance, advisers might offer flattering advice to
strangers out of politeness, provide inflated encouragement to friends
to offer emotional support (Dalal \& Bonaccio, 2010), or use their
recommendations to motivate individuals with particularly ambitious or
competitive personalities. We recognize that multiple motives may
simultaneously drive this behavior, and their relative influence---as
well as advisers' overall tendency to inflate advice---may vary across
situations. However, in this project, we focus specifically on whether
advisers intentionally inflate their advice and explore the potential
consequences of doing so.

One downstream consequence of giving flattering advice is the potential
to reinforce gender differences --- specifically, men are more likely to
receive such inflated advice. Prior work has documented that men receive
more aspirational advice and women more risk-averse advice, which may
explain why women are less likely to apply for and ultimately obtain
more rewarding positions (Kanze et al., 2018). Research on gender
discrimination has focused on the role of unconscious bias (Banaji \&
Greenwald, 1995; Nosek et al., 2009), but interventions that try to
reduce this bias have been largely unsuccessful (Chang et al., 2019;
Paluck \& Green, 2009). Recent work has proposed cognitive, rather than
unconscious, biases that can lead to gender discrimination through the
formation of false beliefs (Hagmann et al., 2024). Here, we propose that
one overlooked mechanism driving gender disparities in career outcomes
may lie in the differential provision of flattering advice, shaped by
the advisee's self-expectations.

Specifically, when advisers aim to make their advice appear flattering,
they may rely less on an objective evaluation of the advisee's
competence and instead anchor their advice on the advisee's
self-expectations. Drawing on the literature on belief-based utility and
expectation-based reference points (Kőszegi \& Rabin, 2006, 2009), we
suggest that advisers may align their advice with how ambitious or
confident the advisee appears, so that they give optimistic but not
entirely unrealistic recommendations. This dynamic may inadvertently
disadvantage women. Research shows that men generally hold higher
self-expectations or are more likely to display confidence, while women
tend to exhibit more modest self-assessments despite comparable
competence (Kanze et al., 2018). As a result, advisers may
unintentionally offer men more ambitious and aspirational advice,
encouraging them to pursue higher-rewarding opportunities, while
offering women more cautious, conservative advice that aligns with their
lower expressed self-expectations. This differential treatment, though
seemingly well-intentioned, can have profound downstream consequences.
By tailoring advice to match the advisee's perceived self-expectations,
advisers may subtly nudge men toward risk-taking, leadership roles, or
competitive opportunities, while directing women toward safer, less
ambitious paths --- ultimately reinforcing existing gender disparities
in career outcomes.

\section{Overview of studies}\label{overview-of-studies}

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their advice to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.\footnote{In some cases, however,
  repeated interaction might also increase incentives for honest
  feedback. Someone who is known to persistently give overoptimistic
  advice may be viewed as less trustworthy in the long run.}

Across all three studies, we use a three-stage design. In Stage 1, we
measure the advisee's performance and their self-expectations. In Stage
2, we show the measured performance to the adviser and ask them to
provide advice on how the advisee can maximize their chances of earning
a bonus. Half of the advisers also receive information about the
advisee's self-expectations. Importantly, this expectation information
is objectively irrelevant to the task, as the bonus is solely based on
the advisee's performance, which is already measured. While in real
life, self-expectations might contain useful insights (such as
unobservable skills or motivation), in our setup, self-expectations
serve no informative value. Moreover, flattering advice provided in this
setup cannot be justified as a motivational boost (like a pep talk)
because the performance has already been fixed. Finally, in Stage 3, we
re-invite the advisees from Stage 1, show them the advice they received,
and observe how they make their final decisions.

In Study 1, we ask advisees to upload photos of themselves and advisers
rank them on their attractiveness. Advisers then recommend which rank
the advisee should bet on, receiving an incentive either for the
advisee's accuracy or for being evaluated favorably. Both groups of
advisers recommend a rank that is more attractive than what they
themselves have evaluated the advisee, and those incentivized for
likability further inflate the ranking. Advisees evaluate advisers more
favorably when they recommend betting on a more attractive rank,
including viewing them as more trustworthy.

In Study 2, we nudge advisees into a preference for competing against
either a group of top-performers or a group of low-performers, through
the use of a default option. Advisers who observe the actual performance
of advisees and therefore have the relevant information needed to make a
recommendation nonetheless take into account the initial preference when
recommending which group they should compete against.

@@

We first examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We
do this by manipulating the incentives for advisers, who receive a bonus
either based on how they are evaluated by an advisee, or a bonus for the
advisee's accuracy.

@@

\caption{\label{fig-study1advice}Participants in Study 2 give advice to
the person they rank 7th most attractive out of 10. When incentivized
for accuracy, the majority recommends betting on that rank. However,
when incentivized for likeability, participants inflate their advice to
flatter the recipient as more attractive.}

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study1advice-1.pdf}}

Next, we turn our attention to the advisers (see
Figure~\ref{fig-study1advice}). In the Accuracy condition, those
uploading selfies are on average advised to bet on rank 6.19. Notably,
this is significantly more attractive than the 7th rank those advisers
had themselves guessed just on the prior screen (\(t(234) = -8.82\),
\(p < .001\)). This suggests that even when incentivized for accuracy,
participants offer flattering advice.\footnote{This shading could be due
  to concerns of avoiding disappointment. However, it could also be that
  advisers are uncertain about the rankings they have given and make a
  recommendation that combines their own belief with a uniform prior.
  Therefore, our analyses focus on the difference between the two
  conditions.} Importantly, and as predicted, we find that advisers in
the Likeable treatment recommend betting on a lower rank, communicating
that they think the participant in the selfie is more attractive (5.38,
\(t(470) = 5.44\), \(p < .001\)). This suggests that advisers infer
flattering someone with pleasant advice would make the adviser appear
more likeable. The distribution shown in Figure~\ref{fig-study1advice}
shows that participants do not simply tell participants that they are
the most attractive person in the group. They may infer that flattering
advise needs to be somewhat realistic to be believable. We return to
this in the general discussion.

{\caption{{When individuals receive advice that implies a high level of
attractiveness (lower rank), they tend to perceive the advice giver as
more likable (Column 1), warm (Column 2). Column 3 shows that advisers
are rated as more trustworthy when they advice lower ranks, but this
relationship is only directional.}{\label{tbl-study1regs}}}\vspace{0pt}
}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]Q[]},
column{1}={halign=l,},
column{2}={halign=c,},
column{3}={halign=c,},
column{4}={halign=c,},
hline{6}={1,2,3,4}{solid, 0.05em, black},
}                     %% tabularray inner close
\toprule
& (1) & (2) & (3) \\ \midrule %% TinyTableHeader
Advised Rank & \num{-0.109}*  & \num{-0.133}** & \num{-0.054}   \\
& (\num{0.046})  & (\num{0.044})  & (\num{0.042})  \\
Constant     & \num{3.777}*** & \num{3.903}*** & \num{3.397}*** \\
& (\num{0.279})  & (\num{0.266})  & (\num{0.257})  \\
N            & \num{146}      & \num{146}      & \num{146}      \\
\bottomrule
\end{talltblr}

\caption{\label{fig-study1rating}Participants tend to rate advisers who
suggest betting on a higher rank (implying greater attractiveness) as
more likable and warm.}

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study1rating-1.pdf}}

\caption{\label{fig-study1ratingByGroup}After participants receive
advice that suggests betting on a rank that was lower than, equal to, or
higher than their own guess, they evaluate advisers based on their
likability and warmth. Participants rate advisers who give flattering
advice and deliver a high evaluation of their attractiveness as more
likable and warm than those whose advice implied a lower evaluation of
their attractiveness.}

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study1ratingByGroup-1.pdf}}

As seen in Figure~\ref{fig-study1rating}, we advisers who suggested that
the advisee was more attractive (lower rank) were indeed rated as more
likeable and warm (b = -0.109, p \textless{} .05; b = -0.133, p
\textless{} .001, respectively; See Columns 1 and 2 of
Table~\ref{tbl-study1regs}). When using their own initial guess as a
reference point, advisers who suggest a worse rank are perceived as less
likable and warm (see Figure~\ref{fig-study1ratingByGroup}).
Interestingly, these benefits are not at the cost of sincerity; advisers
who recommend a more favorable rank are viewed as no less trustworthy
(Column 3 of Table~\ref{tbl-study1regs}).

When advice communicates ego-relevant information (here, people's
attractiveness), we find that the advice people give is contingent on
their incentives. Specifically, when they get rewarded for being more
likeable, they recommend that the advisee bet on a more favorable rank
than when they are incentivized for accuracy. People understand that the
advice they give can influence how others perceive them and often use
flattering advice as a tool to boost their likeablity. Importantly,
advisees do not discount flattering advice and instead evaluate people
who advise them to bet on a more attractive rank as warmer and more
likeable. These gains to interpersonal perceptions do not come at the
cost of trustworthiness, even when the advice is inflated.

We next examine whether advisers take into account the expectations of
advisees when providing a recommendation. A group of participants
(``Advisees'') complete a quiz consisting of ten questions that draw on
ego-relevant domains. They then express a non-binding preference to
compete against a group of high performers or low performers on the same
task, based on their past performance. We nudge participants towards
picking either the low or the high performer group by selecting one of
the two by default. A second group of participants (``Advisers'')
observe Advisees' score on the quiz and, in one treatment, also which
group they had selected. We predicted that Advisees who were nudged
toward the high performer group would be more likely to receive advice
to compete against this high group when Advisers observed their
non-binding choice.

@@

{\caption{{Advice to compete against the High Performer Group in Study
1. Displaying the non-binding choice of those for whom the Low Performer
Group was selected by default makes it less likely that they are advised
to compete against the High Performer Group (Column 1). Column 2
restricts the analysis to advisers who observe advisees' initial
non-binding choice and controls for advisees' score on the quiz.
Standard errors are clustered at the adviser
level.}{\label{tbl-study2regs}}}\vspace{0pt}
}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]},
cell{1}{2}={c=2,}{halign=c,},
column{1}={halign=l,},
column{2}={halign=c,},
column{3}={halign=c,},
hline{13}={1,2,3}{solid, 0.05em, black},
}                     %% tabularray inner close
\toprule
& Advice to Compete Against High Performers &  \\ \cmidrule[lr]{2-3}
& All Advice & Expectations Shown \\ \midrule %% TinyTableHeader
High Default               & \num{0.011}    & \num{0.025}*    \\
& (\num{0.013})  & (\num{0.011})   \\
Expectation Shown          & \num{-0.041}*  &                  \\
& (\num{0.016})  &                  \\
Expectation x High Default & \num{0.037}+   &                  \\
& (\num{0.019})  &                  \\
Score                      &                 & \num{0.168}***  \\
&                 & (\num{0.004})   \\
Constant                   & \num{0.389}*** & \num{-0.503}*** \\
& (\num{0.012})  & (\num{0.020})   \\
N                          & \num{10000}    & \num{5000}      \\
\bottomrule
\end{talltblr}

@@

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument that
reflect higher expectations. We show linear probability models for the
decision of the Advisers, which is equal to 1 if they recommend
competing against the High Performer Group and 0 if they recommend the
Low Performer Group. In Column 1 of Table~\ref{tbl-study2regs}, we show
our main specificiation with the experimental assignment for Advisees
(High Performer Default vs.~Low Performer Default) and Advisers
(Expectation Shown vs.~Expectation Hidden), as well as their
interaction. Because each adviser made ten recommendations, we cluster
standard errors at the adviser level. We find that displaying
expectations of those in the Low Performer Default reduces
recommendations to compete against the Top group by 4 percentage points,
or about 10\% relative to when expectations are hidden. This decrease is
marginally smaller for advisees with a default choice to compete against
the High Performer Group (nudged to have high expectations, though not
necessarily holding them), and whose advice did not differ based on
whether their expectations were shown or hidden. We reported a linear
probability model on advising to compete against the High Performer
group for advisers in the ``Expectations'' treatment, controlling for
the true performance of the advisee in Column 2 of
Table~\ref{tbl-study2regs}. We observe a significant main effect of the
default treatment, where a high default increased the likelihood of
receiving flattering advice. It is worth noting that, since only 65 \%
of advisees in the high default group actually stuck with their default
choice, this represents a conservative test of the effect of high
expectations on the likelihood of receiving flattering advice.

@@

\caption{\label{fig-study2Advice}Advice given in Study 2. Advisees who
were nudged to express high expectations were more likely to receive
advice to compete against the High Performer Group.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-study2Advice-1.pdf}}

@@
